---
title: "30-Day Mortality Rate of Patients with Neoplasm Admitted to MICU"
subtitle: Due Sunday, 3/22 @ 11:59 PM
author: Tony Lim
output:
  # ioslides_presentation: default
  html_document:
    toc: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

2. Introduction. Describe the MIMIC-III data set and what research hypothesis/goal you are to address using this data.

3. Data preparation. Create a study cohort from MIMIC-III corresponding to your research hypothesis/goal. See the examplary code below. Use a CONSORT flow diagram to summarize your steps to create the cohort.

4. Data visualization. Use visualization to summarize the cohort you created. 

5. Analytics. Use at least two analytical approaches to address your research hypothesis/goal. For example, we can use (1) logistic regression and (2) neural network to build a predictive model for the 30-day mortality rate of patients admitted into CCU and compare their predictive performance. Summarize your results in graphs.

6. Conclusions. 

# Introduction
[MIMIC-III](https://mimic.physionet.org/about/mimic/) ((Medical Information Mart for Intensive Care III)) is a free database compromised of over 40,000 patients seen at critical care units of the Beth Israel Deaconess Medical Center between 2001 and 2012. The current version of the database is [v1.4](https://mimic.physionet.org/about/releasenotes/). 

The database includes information such as:

* Demographics
* Vital sign measurements made at the bedside (~1 data point per hour)
* Laboratory test results, procedures
* Medications
* Caregiver notes
* Imaging reports, and
* Mortality (both in and out of hospital)

My research hypothesis predicts that gender will play a significant role in the 30-day mortality rate of patients with neoplasm (cancer) admitted to MICU (Medical Intensive Care Unit). 

MIMIC-III consists of a total of [40 tables](https://mit-lcp.github.io/mimic-schema-spy/index.html). To address the research hypothesis, this report will only utilize:

* [`ADMISSIONS`](https://mimic.physionet.org/mimictables/admissions/)
* [`D_ICD_DIAGNOSES`](https://mimic.physionet.org/mimictables/d_icd_diagnoses/)
* [`DIAGNOSES_ICD`](https://mimic.physionet.org/mimictables/diagnoses_icd/)
* [`DRGCODES`](https://mimic.physionet.org/mimictables/drgcodes/)
* [`PATIENTS`](https://mimic.physionet.org/mimictables/patients/)
* [`TRANSFERS`](https://mimic.physionet.org/mimictables/transfers/)

# Data preparation
Load database libraries and the tidyverse frontend:
```{r}
library(DBI)
library(RPostgreSQL)
library(tidyverse)
library(lubridate)
```

Credentials for using PostgreSQL database. We are going to use username `postgres` with password `postgres` to access the `mimic` database in the schemee `mimiciii`. 
```{r}
# Load configuration settings
dbdriver <- 'PostgreSQL'
#host  <- '127.0.0.1'
#port  <- '5432'
user  <- 'postgres'
password <- 'postgres'
dbname <- 'mimic'
schema <- 'mimiciii'
# Connect to the database using the configuration settings
con <- dbConnect(RPostgreSQL::PostgreSQL(), 
                 dbname = dbname, 
                 #host = host, 
                 #port = port, 
                 user = user, 
                 password = password)
# Set the default schema
dbExecute(con, paste("SET search_path TO ", schema, sep=" "))
con
```

To connect to the `PATIENTS` table in the database:
```{r}
patients <- tbl(con, "patients")
patients %>% print(width = Inf)
```

## Query and subsetting
In this section, we demo how to create a cohort of patients who were directly admitted into MICU and were diagnosed with neoplasm (cancer).

First, we create a (query) table of patients who were directly admitted into MICU.
```{r}
tbl(con, "transfers") %>%
  select(subject_id, hadm_id, prev_careunit, curr_careunit) %>%
  filter(is.na(prev_careunit) & curr_careunit == "MICU") %>%
  select(subject_id, hadm_id) %>%
  distinct() %>%
  print() -> micu_admissions
```

Now we want to restrict to patients with cancer. To find all possible ICD-9 codes related to cancer, we search for string `neoplasm` in the `long_title` of table `D_ICD_DIAGNOSES`:
```{r}
tbl(con, "d_icd_diagnoses") %>%
  filter(str_detect(tolower(long_title), "neoplasm")) %>%
  print() -> neo_codes
```

`DIAGNOSES_ICD` table stores the diagnosis of each admission. We use `semi_join()` to keep the rows in `DIAGNOSES_ICD` that match the ICD-9 codes related to cancer:
```{r}
tbl(con, "diagnoses_icd") %>%
  semi_join(neo_codes, by = "icd9_code") %>% 
  print() -> neo_admissions
```

Cancer may not be listed as the principal diagnosis; as explained in [the documentation for the `patients` table](https://mimic.physionet.org/mimictables/diagnoses_icd/), the `seq_num` field is a priority ranking for the diagnoses generated at the end of stay. In order to focus on patients for whom cancer was central to their hospitalization, we will include records with neoplasm in any of the first five diagnosis positions, according to the `seq_num` field. To avoid duplicate admissions, we use `group_by()` and `top_n()` to limit the query to the first cancer diagnosis for each admission.
```{r}
neo_admissions %>%
  filter(seq_num <= 5) %>%
  group_by(subject_id, hadm_id) %>%
  filter(min_rank(seq_num) <= 1) %>%
  ungroup() %>%
  select(subject_id, hadm_id, icd9_code, seq_num) %>% 
  print() -> neo_admissions
```

Now we `inner_join` the table of admissions to MICU and the table of admissions that include neoplasm diagnosis.
```{r}
micu_admissions %>%
  inner_join(neo_admissions, by = c("subject_id", "hadm_id")) %>% 
  print() -> study_admissions
```

## Transform and augment query tables
Now we create a logical variable indicating whether cancer is the principal diagonosis or not (according to `seq_num`).
```{r}
study_admissions %>%
  mutate(principal_dx = seq_num == 1) %>%
  select(-seq_num) %>%
  print() -> study_admissions
```

We want to add information about the severity of patients’ ailments. The `DRGCODES` table contains, for `DRG` codes from the All Payers Registry (APR), severity and mortality indicators. We pull the drug severity information and right-join it to our query table.
```{r}
tbl(con, "drgcodes") %>%
  filter(str_detect(drg_type, "APR")) %>%
  select(subject_id, hadm_id, drg_severity) %>%
  right_join(study_admissions, by = c("subject_id", "hadm_id")) %>%
  mutate(drg_severity = ifelse(is.na(drg_severity), 1, drg_severity)) %>%
  print() -> study_admissions
```

Pull the admission time `admittime`, discharge time `dischtime`, date of birth `dob`, and date of death `dod`. We are interested in the  mortaility rate 30 days after discharge. So we only keep patients who didn't die in hospital.
```{r}
study_admissions %>%
  left_join(
    select(tbl(con, "admissions"),
           subject_id, hadm_id, admittime, dischtime, hospital_expire_flag
    ), by = c("subject_id", "hadm_id")
  ) %>%
  filter(hospital_expire_flag == 0) %>% # patients who did not die in hospital
  select(-hospital_expire_flag) %>%
  left_join(
    select(tbl(con, "patients"), subject_id, dob, dod),
    by = "subject_id"
  ) %>%
  print(width = Inf) -> study_admissions
```

To add `age` (at admission) variable into the table. [The documentation for the patients table](https://mimic.physionet.org/mimictables/patients/) explains that patients of 90 years and older had their ages artificially inflated, so we remove these patients from the analysis.
```{r}
study_admissions %>%
  mutate(tt_death = date_part("day", dod) - date_part("day", dischtime)) %>%
  mutate(mortality = tt_death <= 30) %>%
  mutate(age = date_part("year", admittime) - date_part("year", dob)) %>%
  filter(age < 90) %>%
  mutate(age = age - ifelse(
    date_part("month", admittime) < date_part("month", dob) |
      (
        date_part("month", admittime) == date_part("month", dob) &
          date_part("day", admittime) < date_part("day", dob)
      ),
    1,
    0
  )) %>%
  select(-admittime, -dischtime, -dob, -dod, -tt_death) %>%
  select(subject_id, hadm_id, age, mortality, everything()) %>%
  print() -> study_admissions
```

Many mortality indicators are missing, due to neither the hospital database nor the social security database having a record of these patients’ deaths. We could convert these to `FALSE` values, but it may be helpful to retain in the analytic table this information on whether deaths were recorded at all, e.g. for validation or sensitivity testing.

Finally, let's merge some demographic information (ethnicity, gender) into our study `study_admissions`.
```{r}
tbl(con, "admissions") %>%
  select(subject_id, ethnicity) %>%
  distinct() %>%
  print() -> study_subjects
```
```{r}
tbl(con, "patients") %>%
  select(subject_id, gender) %>%
  distinct() %>%
  full_join(study_subjects, by = "subject_id") %>%
  print() -> study_subjects
```
```{r}
study_subjects %>%
  semi_join(study_admissions, by = "subject_id") %>%
  print() -> study_subjects
```

Let's resolve some diversity and inconsistency in the `ethnicity` field:
```{r}
unknown_ethnicity <- c(
  "OTHER",
  "UNABLE TO OBTAIN",
  "UNKNOWN/NOT SPECIFIED",
  "MULTI RACE ETHNICITY",
  "PATIENT DECLINED TO ANSWER",
  "UNKNOWN"
)

study_subjects %>%
  collect() %>%
  mutate(ethnic_group = case_when(
    str_detect(ethnicity, "^ASIAN") ~ "ASIAN",
    str_detect(ethnicity, "^BLACK") ~ "BLACK",
    str_detect(ethnicity, "^HISPANIC") ~ "HISPANIC",
    str_detect(ethnicity, "^WHITE") ~ "WHITE",
    ethnicity %in% unknown_ethnicity ~ NA_character_,
    TRUE ~ NA_character_
  )) %>%
  select(subject_id, gender, ethnic_group) %>%
  print() -> study_subjects
```

Some patients are coded as belonging to more than one ethnic group. To resolve these inconsistencies, we define a helper function to pick the modal value from a vector of values in R, which can be used by the `summarize()` function to choose one ethnic group for each patient.
```{r}
most <- function(x) {
  if (all(is.na(x))) return(NA_character_)
  y <- table(x, useNA = "no")
  if (length(which(y == max(y))) > 1) return(NA_character_)
  return(names(y)[which.max(y)])
}

study_subjects %>%
  group_by(subject_id) %>%
  summarize(ethnic_group = most(ethnic_group)) %>%
  ungroup() %>%
  mutate(ethnic_group = ifelse(is.na(ethnic_group), "UNKNOWN", ethnic_group)) %>%
  print() -> subject_ethnic_groups
```
```{r}
study_subjects %>%
  select(subject_id, gender) %>%
  left_join(subject_ethnic_groups, by = "subject_id") %>%
  print() -> study_subjects
```

Now we add the demographic information `gender` and `ethnicity` into our `study_admissions` table:
```{r}
study_admissions %>%
  left_join(study_subjects, by = "subject_id", copy = TRUE) %>%
  print() -> study_admissions
```

## CONSORT Flow Diagrams
The following [CONSORT Flow Diagrams](http://www.consort-statement.org/consort-statement/flow-diagram) plot the flow of data selection of the patient cohort. 
```{r plot}
library(shape)
library(diagram)

# set margins and multiplot
par(mfrow = c(1, 1))
par(mar = c(0, 0, 0, 0))

# initialise a plot device
openplotmat()

# position of boxes
# 1st column indicates x axis position between 0 and 1
# 2nd column indicates y axis position between 0 and 1
# automatically assigns vertical position
num_of_boxes <- 4
auto_coords = coordinates(num_of_boxes)
vert_pos = rev(auto_coords[,1])
box_pos <- matrix(nrow = num_of_boxes, ncol = 2, data = 0)
box_pos[1,] = c(0.20, vert_pos[1]) # 1st box
box_pos[2,] = c(0.70, vert_pos[2]) # 2nd box
box_pos[3,] = c(0.70, vert_pos[3]) # 3rd box
box_pos[4,] = c(0.20, vert_pos[4]) # etc...

# content of boxes
box_content <- matrix(nrow = num_of_boxes, ncol = 1, data = 0)
box_content[1] = "All diagnoses in DIAGNOSES_ICD \n n = 651,047" # 1st box
box_content[2] = "Exclude those without 'neoplasm' \n n = 632,362" # 2nd box
box_content[3] = "Exclude if 'neoplasm' isn't in first five seq_num
  n = 12,947" # 3rd box
box_content[4] = "neo_admissions \n n = 5,738"

# adjust the size of boxes to fit content
box_x <- c(0.20, 0.25, 0.25, 0.20)
box_y <- c(0.07, 0.07, 0.07, 0.07)

# Draw the arrows
straightarrow(from = c(box_pos[1,1],box_pos[2,2]), to = box_pos[2,], lwd = 1)  
straightarrow(from = c(box_pos[1,1],box_pos[3,2]), to = box_pos[3,], lwd = 1)  
straightarrow(from = c(box_pos[1,1],box_pos[4,2]), to = box_pos[4,], lwd = 1)  
straightarrow(from = box_pos[1,], to = box_pos[4,], lwd = 1)

# Draw the boxes
for (i in 1:num_of_boxes) {
  textrect(mid = box_pos[i,], radx = box_x[i], rady = box_y[i], 
           lab = box_content[i], 
           shadow.col = "grey")
  }
```

```{r}
# set margins and multiplot
par(mfrow = c(1, 1))
par(mar = c(0, 0, 0, 0))

# initialise a plot device
openplotmat()

# position of boxes
# 1st column indicates x axis position between 0 and 1
# 2nd column indicates y axis position between 0 and 1
# automatically assigns vertical position
num_of_boxes <- 4
auto_coords = coordinates(num_of_boxes)
vert_pos = rev(auto_coords[,1])
box_pos <- matrix(nrow = num_of_boxes, ncol = 2, data = 0)
box_pos[1,] = c(0.20, vert_pos[1]) # 1st box
box_pos[2,] = c(0.70, vert_pos[2]) # 2nd box
box_pos[3,] = c(0.70, vert_pos[3]) # 3rd box
box_pos[4,] = c(0.20, vert_pos[4]) # etc...

# content of boxes
box_content <- matrix(nrow = num_of_boxes, ncol = 1, data = 0)
box_content[1] = "All admissions in TRANSFER \n n = 261,897"
box_content[2] = "Exclude admissions if prev_careunit = NA or
  curr_careunit != MICU \n n = 241,604"
box_content[3] = "Exclude admissions not in neo_admissions \n n = 17,256" 
box_content[4] = "Study cohort \n n = 3,037"

# adjust the size of boxes to fit content
box_x <- c(0.20, 0.25, 0.25, 0.20)
box_y <- c(0.07, 0.10, 0.10, 0.07)

# Draw the arrows
straightarrow(from = c(box_pos[1,1],box_pos[2,2]), to = box_pos[2,], lwd = 1)  
straightarrow(from = c(box_pos[1,1],box_pos[3,2]), to = box_pos[3,], lwd = 1)  
straightarrow(from = c(box_pos[1,1],box_pos[4,2]), to = box_pos[4,], lwd = 1)  
straightarrow(from = box_pos[1,], to = box_pos[4,], lwd = 1)

# Draw the boxes
for (i in 1:num_of_boxes) {
  textrect(mid = box_pos[i,], radx = box_x[i], rady = box_y[i], 
           lab = box_content[i], 
           shadow.col = "grey")
  }
```

# Data visualization
## Age
```{r, warning=FALSE, message=FALSE}
study_admissions %>%
  arrange(subject_id, desc(age)) %>%
  group_by(subject_id) %>%
  filter(row_number() == 1) %>%
  ggplot +
  geom_histogram(mapping = aes(x = age)) +
  labs(title = "Age distribution",
       subtitle = "n = 1,697",
       x = "Age",
       y = "Count")
```

## Gender
```{r}
study_admissions %>%
  arrange(subject_id) %>%
  group_by(subject_id, gender) %>%
  filter(row_number() == 1) %>%
  ggplot +
  geom_bar(mapping = aes(x = gender)) +
  labs(title = "Gender distribution",
       subtitle = "n = 1,697",
       x = "Gender",
       y = "Count")
```

## Ethnicity
```{r}
study_admissions %>%
  arrange(subject_id) %>%
  group_by(subject_id, ethnic_group) %>%
  filter(row_number() == 1) %>% 
  ggplot +
  geom_bar(mapping = aes(x = ethnic_group)) + 
  labs(title = "Ethnicity distribution",
       subtitle = "n = 1,697",
       x = "Ethnicity", 
       y = "Count")
```

## DRG severity
```{r}
study_admissions %>%
  arrange(subject_id, desc(drg_severity)) %>%
  group_by(subject_id, drg_severity) %>%
  filter(row_number() == 1) %>%
  ggplot +
  geom_bar(mapping = aes(x = drg_severity, fill = gender)) +
  labs(title = "DRG severity by gender",
       subtitle = "n = 1,801",
       x = "DRG severity",
       y = "Count")
```

## Principal diagnosis
```{r}
study_admissions %>%
  arrange(subject_id, principal_dx) %>%
  group_by(subject_id, principal_dx) %>%
  filter(row_number() == 1) %>%
  ggplot +
  geom_bar(mapping = aes(x = principal_dx, fill = gender)) +
  labs(title = "Principal diagnosis by gender",
       subtitle = "n = 1,746",
       x = "Principal diagnosis",
       y = "Count")
```

## Mortality
```{r}
study_admissions %>%
  arrange(subject_id, mortality) %>%
  group_by(subject_id, mortality) %>%
  filter(row_number() == 1) %>%
  ggplot +
  geom_bar(mapping = aes(x = mortality, fill = gender)) +
  labs(title = "Mortality by gender",
       subtitle = "n = 1,697",
       x = "Mortality",
       y = "Count")
```

# Data analysis
## Creating training and testing samples
```{r}
df <- as.data.frame(study_admissions)
df <- replace_na(df, list(mortality = FALSE))

table(df$mortality)

set.seed(203)
sample <- sample(c(TRUE, FALSE), nrow(df), replace = T, prob = c(0.6, 0.4))

train <- df[sample, ]
test <- df[!sample, ]
```

## Logistic regression
```{r}
glm(mortality ~ age + drg_severity + gender, data = train, 
                  family = binomial(link = "logit"), maxit = 100) %>%
  summary()

logit_mod <- glm(mortality ~ age + drg_severity + gender, data = train, 
                  family = binomial(link = "logit"), maxit = 100)

predicted <- plogis(predict(logit_mod, test))

df
```

## Neural network
```{r}
library(keras)

mx <- as.matrix(as.data.frame(study_admissions))


```

## Close the connection to a database
Close the connection:
```{r}
# dbDisconnect(con)
```

# Conclusion

